An Empirical Study of Invasion and Resistance for Iterated Prisoner's Dilemma Strategies

### 2017-05-09 main.tex

% TODO Add discussion about ZD strategies and discuss how this corresponds to
% case of N = 2. Include pointing at \cite{Adami2013} which discusses how ZD
% strategies are not fantastic in population dynamics.
% Add \cite{Li2014}

@@

\subsection{The Axelrod Library}

@@

\begin{enumerate}
    \item The top strategy is an extortionate Zero determinant
        strategy~\cite{Press2012} with parameters \(l=1\) and \(s=1/4\).
    \item The Collective strategy has a simple handshake mechanism (a
        cooperation followed by a defection on the first move). As long as the
        opponent plays the same handshake and does not defect in the future it
        cooperates. Otherwise it defects for all rounds~\cite{Li2009}.  This
        strategy was specifically designed for evolutionary processes.
    \item The finite state machine strategy % TODO Perhaps refer to earlier
        % discussion in the methodology section about these FSM strategies
    \item The Feld strategy is the corresponding strategy submitted to Axelrod's
        first tournament~\cite{Axelrod1980a}: it punishes defections but
        otherwise defects with a random probability that decays over time.
    \item The final strategy in the top five is another extortionate Zero
        determinant strategy~\cite{Press2012} with parameters \(l=p\).
\end{enumerate}

@@

In this section we focus on the ability of a mutant strategy to invade: the
probability of 1 individual of a given type successfully becoming fixated in
a population of \(N - 1\) other individuals, denoted by \(x_i\).
Figures~\ref{fig:fixation_heatmap_invade} shows these values for
the players along the vertical axis when matched against the players on the
horizontal axis. For population size $N$, a neutral mutant has fixation
probability $x_1 = 1 / N$. This information is summarised in Figure~\ref{fig:fixation_boxplot_invade}
showing the median fixation as well as the neutral fixation for each given
scenario.

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=.8\textwidth]{./img/fixation_heatmap_3_invade.pdf}
        \caption{\(N=3\)}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=.8\textwidth]{./img/fixation_heatmap_7_invade.pdf}
        \caption{\(N=7\)}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=.8\textwidth]{./img/fixation_heatmap_12_invade.pdf}
        \caption{\(N=12\)}

    %TODO Replace N=12 with N=14
    \end{subfigure}%
    %TODO Remove these? I don't feel they are effective

    \caption{Pairwise fixation probability \(x_{1}\) of all strategies}
    \label{fig:fixation_heatmap_invade}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./img/boxplot_3_invade.pdf}
        \caption{\(N=3\)}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./img/boxplot_7_invade.pdf}
        \caption{\(N=7\)}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./img/boxplot_12_invade.pdf}
        \caption{\(N=12\)}
    \end{subfigure}%

    % TODO Replace N=12 with N=14 (when we have more data).
    \caption{Median probabilities \(x_{1}\) of all strategies as well as the
    neutral fixation probability}
    \label{fig:fixation_boxplot_invade}
\end{figure}

For \(N\in\{3, 7, 12\}\) the top five strategies are given in
% TODO Change N=12 to N=14
Tables~\ref{tbl:top_five_invade}.

\begin{table}[!hbtp]
    \begin{subfigure}[t]{\textwidth}
        \centering
        \input{./tex/summary_top_3_invade.tex}
        \caption{\(N=3\)}
    \end{subfigure}
    \begin{subfigure}[t]{\textwidth}
        \centering
        \input{./tex/summary_top_7_invade.tex}
        \caption{\(N=7\)}
    \end{subfigure}
    \begin{subfigure}[t]{\textwidth}
        \centering
        \input{./tex/summary_top_12_invade.tex}
        \caption{\(N=12\)}
    \end{subfigure}
    %TODO Change N=12 to N=14
    \caption{Properties of top five invaders}
    \label{tbl:top_five_invade}
\end{table}

It can be seen that apart from the Collective Strategy, none of the strategies
of Table~\ref{tbl:summary_top_2} perform well for \(N\in\{3, 7, 12\}\). The new
% TODO Change N=12 to N=14
high performing strategies are:

\begin{itemize}
    \item Predator, a finite state machine described in~\cite{Ashlock2006}.
    \item Prober 4, complex strategy with an initial 20 move sequence of
        cooperations and defections~\cite{prison}. This initial sequence serves
        as approximate handshake.
    \item Remorseful Prober, a strategy that will not immediately retaliate when
        it recognises that the opponent is retaliating to a random
        defection~\cite{Li2011}.
    \item  Worse and Worse 2: plays Tit For Tat for 20 moves and then defects
        with with growing probability~\cite{prison}.
    \item Tester: a strategy submitted to the second of Axelrod's tournaments
        \cite{Axelrod1980b}.
        % TODO Check if this list is still correct for N=14
\end{itemize}

As well as noting that the memory length and complexity of these strategies are
much greater than one, it is interesting to note that none of them are akin to
memory one strategies. Most are not stochastic.

@@

Figures~\ref{fig:fixation_heatmap_resist} show \(x_{N-1}\)
the players along the vertical axis when matched against the players on the
horizontal axis. This information is summarised in Figure~\ref{fig:fixation_boxplot_resist}
showing the median fixation as well as the neutral fixation for each given
scenario.

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=.8\textwidth]{./img/fixation_heatmap_3_resist.pdf}
        \caption{\(N=3\)}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=.8\textwidth]{./img/fixation_heatmap_7_resist.pdf}
        \caption{\(N=7\)}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=.8\textwidth]{./img/fixation_heatmap_12_resist.pdf}
        \caption{\(N=12\)}

    %TODO Replace N=12 with N=14
    \end{subfigure}%
    %TODO Remove these? I don't feel they are effective

    \caption{Pairwise fixation probability \(x_{N-1}\) of all strategies}
    \label{fig:fixation_heatmap_resist}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./img/boxplot_3_resist.pdf}
        \caption{\(N=2\)}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./img/boxplot_7_resist.pdf}
        \caption{\(N=7\)}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./img/boxplot_12_resist.pdf}
        \caption{\(N=12\)}
    \end{subfigure}%

    % TODO Replace N=12 with N=14 (when we have more data).
    \caption{Median probabilities \(x_{N-1}\) of all strategies as well as the
    neutral fixation probability}
    \label{fig:fixation_boxplot_resist}
\end{figure}

Table~\ref{tbl:top_five_resist} shows the top five strategies when ranked
according to \(x_{N-1}\) for \(N\in\{3, 7, 12\}\).
% TODO Change 12 to 14
Once again none of the short memory strategies from
Section~\ref{sec:two_individuals} perform well for high \(N\).

Three strategies have both high \(x_1\) and high \(x_{N-1}\):

\begin{itemize}
    \item Collective
    \item Predator
    \item Prober 4
\end{itemize}

However, Remorseful Prober, Worse and Worse 2 and Tester no longer do as well.
There are two strategies that are only top performers in \(x_{N-1}\):

\begin{itemize}
    \item Handshake: a slightly less aggressive version of the Collective
        strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself: just as the Collective strategy does. However it
        is not aggressive enough to invade other populations.
    \item Winner 21: a strategy that makes its decision deterministically based
        on 1 round of its own strategy and 2 of the opponent's strategy
        \cite{Mathieu2015}.
\end{itemize}
% TODO Revise this if need be for N=14

@@

\begin{table}[!hbtp]
    \begin{subfigure}[t]{\textwidth}
        \centering
        \input{./tex/summary_top_3_resist.tex}
        \caption{\(N=3\)}
    \end{subfigure}
    \begin{subfigure}[t]{\textwidth}
        \centering
        \input{./tex/summary_top_7_resist.tex}
        \caption{\(N=7\)}
    \end{subfigure}
    \begin{subfigure}[t]{\textwidth}
        \centering
        \input{./tex/summary_top_12_resist.tex}
        \caption{\(N=12\)}
    \end{subfigure}
    %TODO Change N=12 to N=14
    \caption{Properties of top five resistors}
    \label{tbl:top_five_resist}
\end{table}

@@

Tables~\ref{tbl:ranks_v_size_invade},~\ref{tbl:ranks_v_size_resist}
and~\ref{tbl:ranks_v_size_coexist} show the same information for the strategies
that rated high for \(N=2\) and \(N=12\).
% TODO Change to N=14?

@@

Tables~\ref{tbl:correlation_coeficients_invade},~\ref{tbl:correlation_coeficients_resist}
and~\ref{tbl:correlation_coeficients_coexist} show the correlation coefficients
of the ranks in of strategies in differing population size. This is shown
graphically in Figure~\ref{fig:correlation_coefficients}. It is immediate to
note that how well a strategy performs in any Moran process for \(N>2\) has
little to do with the performance for \(N=2\). This illustrates why the strong
performance of zero determinant strategies predicted in \cite{Press2012} does
not extend to larger populations. This was discussed theoretically in
\cite{Adami2013} however not observed empirically at the scale presented here.

@@

For \(N=2\), memory one strategies perform well, in particular as predicted by
\cite{Press2012} zero determinant strategies rank highly. However, there are no
memory one strategies in the top 5 performing strategies for \(N>3\). This is
due to their lack of sophistication which allows them to recognise and adjust to
their opponent.

>>>>

@@

@@

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism (a cooperation followed by a defection on the first
        move). As long as the opponent plays the same handshake and does not
        defect in the future it cooperates. Otherwise it defects for all
        rounds~\cite{Li2009}. This strategy was specifically designed for
        evolutionary processes.
    \item The finite state machine strategy % TODO Perhaps refer to earlier
        % discussion in the methodology section about these FSM strategies
    \item The Defector: it always defects. As it has little potential
        interaction with itself, recall that \(N=2\) is considered, its
        aggressiveness is rewarded.
    \item The Aggravater strategy which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in~\cite{Ashlock2006}.
\end{enumerate}

@@

In this section we focus on the ability of a mutant strategy to invade: the
probability of 1 individual of a given type successfully becoming fixated in a
population of \(N - 1\) other individuals, denoted by \(x_i\). The fixation
probabilities are shown in
Figures~\ref{fig:boxplot_3_invade},~\ref{fig:boxplot_7_invade}
and~\ref{fig:boxplot_14_invade} for \(N\in\{3, 7, 14\}\) showing the mean
fixation as well as the neutral fixation for each given scenario.

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.8\textheight]{./img/boxplot_3_invade.pdf}
    \caption{The fixation probability \(x_1\) for \(N=3\)}
    \label{fig:boxplot_3_invade}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.8\textheight]{./img/boxplot_7_invade.pdf}
    \caption{The fixation probability \(x_1\) for \(N=7\)}
    \label{fig:boxplot_7_invade}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.8\textheight]{./img/boxplot_14_invade.pdf}
    \caption{The fixation probability \(x_1\) for \(N=14\)}
    \label{fig:boxplot_14_invade}
\end{figure}

The top five strategies are given in Tables~\ref{tbl:top_five_invade}.

\begin{table}[!hbtp]
    \begin{subfigure}[t]{\textwidth}
        \centering
        \input{./tex/summary_top_3_invade.tex}
        \caption{\(N=3\)}
    \end{subfigure}
    \begin{subfigure}[t]{\textwidth}
        \centering
        \input{./tex/summary_top_7_invade.tex}
        \caption{\(N=7\)}
    \end{subfigure}
    \begin{subfigure}[t]{\textwidth}
        \centering
        \input{./tex/summary_top_14_invade.tex}
        \caption{\(N=14\)}
    \end{subfigure}
    \caption{Properties of top five invaders}
    \label{tbl:top_five_invade}
\end{table}

It can be seen that apart from CS, none of the strategies
of Table~\ref{tbl:summary_top_2} perform well for \(N\in\{3, 7, 14\}\). The new
high performing strategies are:

\begin{itemize}
    \item Prober 4, complex strategy with an initial 20 move sequence of
        cooperations and defections~\cite{prison}. This initial sequence serves
        as approximate handshake.
    \item Grudger (which only performs well for \(N=3\)), starts by cooperating
        but will defect if at any point the opponent has defected.
    \item MEM2, an infinite memory strategy that switches between TfT, Tf2T, and
        Defector~\cite{Li2014}.
    \item Fool Me Once, forgives one defection but defects forever once a second
        defection takes place.
    \item  PSO Gambler and Evolved Lookerup 2 2 2: are strategies that make use
        of a lookup table mapping the first 2 moves of the opponent as well as
        the last 2 moves of both players to an action. The PSO gambler is a
        stochastic version which maps those states to probabilities of
        cooperating. The lookerup was described in \cite{Knight2016}.
	\item The evolved ANN strategies are neural networks that map a number of
		attributes (first move, number of cooperations, last move etc\dots) to
		an action. Both of these have been trained using an evolutionary
		algorithm and the ANN 5 was trained to perform well in a noisy
		tournament.
	\item The Evolved FSM 16, is a 16 state finite machine strategy that has
		been trained using an evolutionary algorithm.
\end{itemize}

As well as noting that the memory length and complexity of these strategies are
much greater than one, it is interesting to note that none of them are akin to
memory one strategies. Only one is stochastic.

@@

The fixation
probabilities are shown in
Figures~\ref{fig:boxplot_3_resist},~\ref{fig:boxplot_7_invade}
and~\ref{fig:boxplot_14_resist} for \(N\in\{3, 7, 14\}\) showing the mean
fixation as well as the neutral fixation for each given scenario.

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.8\textheight]{./img/boxplot_3_resist.pdf}
    \caption{The fixation probability \(x_{N-1}\) for \(N=3\)}
    \label{fig:boxplot_3_resist}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.8\textheight]{./img/boxplot_7_resist.pdf}
    \caption{The fixation probability \(x_{N-1}\) for \(N=7\)}
    \label{fig:boxplot_6_resist}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.8\textheight]{./img/boxplot_14_resist.pdf}
    \caption{The fixation probability \(x_{N-1}\) for \(N=14\)}
    \label{fig:boxplot_14_resist}
\end{figure}

Table~\ref{tbl:top_five_resist} shows the top five strategies when ranked
according to \(x_{N-1}\) for \(N\in\{3, 7, 14\}\).
Once again none of the short memory strategies from
Section~\ref{sec:two_individuals} perform well for high \(N\).

\begin{table}[!hbtp]
    \begin{subfigure}[t]{\textwidth}
        \centering
        \input{./tex/summary_top_3_resist.tex}
        \caption{\(N=3\)}
    \end{subfigure}
    \begin{subfigure}[t]{\textwidth}
        \centering
        \input{./tex/summary_top_7_resist.tex}
        \caption{\(N=7\)}
    \end{subfigure}
    \begin{subfigure}[t]{\textwidth}
        \centering
        \input{./tex/summary_top_14_resist.tex}
        \caption{\(N=14\)}
    \end{subfigure}
    \caption{Properties of top five resistors}
    \label{tbl:top_five_resist}
\end{table}

There are 3 strategies that are only top performers in \(x_{N-1}\):

\begin{itemize}
    \item Handshake: a slightly less aggressive version of the Collective
        strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself: just as the Collective strategy does. However it
        is not aggressive enough to invade other populations.
    \item Winner 21: a strategy that makes its decision deterministically based
        on 1 round of its own strategy and 2 of the opponent's strategy
        \cite{Mathieu2015}.
	\item Predator: a finite state machine described in~\cite{Ashlock2006}
\end{itemize}

@@

@@

Tables~\ref{tbl:ranks_v_size_invade},~\ref{tbl:ranks_v_size_resist}
and~\ref{tbl:ranks_v_size_coexist} show the same information for the strategies
that rated high for \(N=2\) and \(N=14\).

@@

Tables~\ref{tbl:correlation_coeficients_invade},~\ref{tbl:correlation_coeficients_resist}
and~\ref{tbl:correlation_coeficients_coexist} show the correlation coefficients
of the ranks of strategies in differing population size. This is shown
graphically in Figure~\ref{fig:correlation_coefficients}. It is immediate to
note that how well a strategy performs in any Moran process for \(N>2\) has
little to do with the performance for \(N=2\). This illustrates why the strong
performance of zero determinant strategies predicted in \cite{Press2012} does
not extend to larger populations. This was discussed theoretically in
\cite{Adami2013} however not observed empirically at the scale presented here.

@@

Memory one strategies do not perform well, as predicted by \cite{Press2012}.
However, there are no memory one strategies in the top 5 performing strategies
for \(N>3\). This is due to their lack of sophistication which allows them to
recognise and adjust to their opponent. Some very sophisticated strategies
proves to be high performers for invasion: these are infinite memory strategies
which have been trained using a number of reinforcement learning algorithms.
Interestingly they have been trained to perform well in tournaments and not
Moran processes which highlights the potentially for improvement.
